---
title: "Biostat 203B Homework 4"
subtitle: Due Mar 22 @ 11:59PM
author: Rui Xu
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task

In this assignment, you are to write a report analyzing the electronic health record (EHR) data MIMIC-III. You report will demostrate your knowledge of working with PostgreSQL database, data visualization, and commonly used analytical methods such as logistic regression and neural network. Your report should include at least following parts:  

1. An informative title. For example, _30-Day Mortality Rate of Myocardia Infarction Patients Admitted to CCU_.  

2. Introduction. Describe the MIMIC-III data set and what research hypothesis/goal you are to address using this data.

3. Data preparation. Create a study cohort from MIMIC-III corresponding to your research hypothesis/goal. See the examplary code below. Use a CONSORT flow diagram to summarize your steps to create the cohort.

4. Data visualization. Use visualization to summarize the cohort you created. 

5. Analytics. Use at least two analytical approaches to address your research hypothesis/goal. For example, we can use (1) logistic regression and (2) neural network to build a predictive model for the 30-day mortality rate of patients admitted into CCU and compare their predictive performance. Summarize your results in graphs.

6. Conclusions. 

# Learning resources about analyzing EHR data

- _Secondary Analysis of Electronic Health Records_: <https://link.springer.com/book/10.1007/978-3-319-43742-2> 

- _The Book of OHDSI_: <https://ohdsi.github.io/TheBookOfOhdsi/>. 

- The GitHub repository <https://github.com/MIT-LCP/mimic-code> contains some code examples for working with the MIMIC-III data. Following sample code derives from <https://github.com/MIT-LCP/mimic-code/blob/master/tutorials/dplyr-frontend/intro.md>. 

# Title: _More-than-5-Day ICU Stay Rate of Pregnant Woman_

# Introduction

The MIMIC-III (*M*edical *I*nformation *M*art for *I*ntensive *C*are III) is a database comprising deidentified health-related data associated with over forty thousand patients who stayed in critical care units of the Beth Israel Deaconess Medical Center between 2001 and 2012. The database includes information such as admission information, patients' demographics, diagnose information, and ICU stay information.
I plan to use the information to construct a study cohort, where patients are dignosed complications of pregancy, childbirth, and the puerperium, and admitted into ICU. The study cohort covers the age, ethnicity of patients, the severity of ailments, ICD-9 code, principal or not diagnose information, and the length of stay in ICU. Two models are built so that whether a pregant patient stay in ICU for more than 5 days can be predicted with the first 5 variable information.

# Data preparation

## Connect to PostgresSQL database

Load database libraries and the tidyverse frontend:
```{r}
library(DBI)
library(RPostgreSQL)
library(tidyverse)
library(lubridate)
```

Credentials for using PostgreSQL database. We are going to use username `postgres` with password `postgres` to access the `mimic` database in the schemee `mimiciii`. 
```{r}
# Load configuration settings
dbdriver <- 'PostgreSQL'
#host  <- '127.0.0.1'
#port  <- '5432'
user  <- 'postgres'
password <- 'postgres'
dbname <- 'mimic'
schema <- 'mimiciii'
# Connect to the database using the configuration settings
con <- dbConnect(RPostgreSQL::PostgreSQL(), 
                 dbname = dbname, 
                 #host = host, 
                 #port = port, 
                 user = user, 
                 password = password)
# Set the default schema
dbExecute(con, paste("SET search_path TO ", schema, sep=" "))
con
```
List tables in the `mimic` database:
```{r}
dbListTables(con)
```

## Query and subsetting

In this section, we demo how to create a cohort of patients who were admitted into ICU and were diagnosed with pregnancy.

First we create a (query) table of patients who were admitted into ICU.
```{r}
tbl(con, "icustays") %>%
  select(subject_id, hadm_id, los) %>%
  filter(!is.na(los)) %>% 
  # group_by(subject_id) %>% 
  # mutate(los = sum(los)) %>% 
  # filter(los >= 5) %>% 
  print() -> icustays
```

Now we want to restrict to complications of pregancy, childbirth, and the puerperium. To find all possible ICD-9 codes related to this, we search for the list of ICD-9 codes 63000-67999 in the `icd9_code` of table `d_icd_diagnoses`:
```{r}
code <- seq(from = 630, to = 679) %>%
  append(seq(from = 63000, to = 67999))
# code <- paste(c("^"), 630:679, sep = "")
code <- as.character(code) #%>%
#   str_c(collapse = "|")
# code <- paste("(", code, ")", sep = "")
tmp <- tbl(con, "d_icd_diagnoses")
icd <- tribble(~icd9_code, ~long_title)
for (c in code){
  icd9 <- tmp %>%
    filter(icd9_code == c) %>% 
    select(icd9_code, long_title) %>% as.tibble
  icd <- bind_rows(icd, icd9)
  # mutate(icd9_code = as.integer(icd9_code)) %>% 
  # filter(icd9_code >= 63000, icd9_code < 68000) %>%
}
  icd %>% print()
  dbWriteTable(con, "icd", icd, overwrite = TRUE)
```

`diagnoses_icd` table stores the diagnosis of each admission. We use `semi_join()` to keep the rows in `diagnoses_icd` that match the ICD-9 codes related to complications of pregancy, childbirth, and the puerperium:
```{r}
tbl(con, "diagnoses_icd") %>%
  semi_join(icd, by = "icd9_code", copy = TRUE) %>%
  print() -> diagnose
```

MI may not be listed as the principal diagnosis; as explained in [the documentation for the `patients` table](https://mimic.physionet.org/mimictables/diagnoses_icd/), the `seq_num` field is a priority ranking for the diagnoses generated at the end of stay. In order to focus on patients for whom MI was central to their hospitalization, we will include records with MI in any of the first five diagnosis positions, according to the `seq_num` field. To avoid duplicate admissions, we use `group_by()` and `top_n()` to limit the query to the first MI diagnosis for each admission.
```{r}
diagnose %>%
  filter(seq_num <= 5) %>%
  group_by(subject_id, hadm_id) %>%
  # top_n(1, wt = seq_num) %>% #  not working. bug? use following as workaround
  filter(min_rank(seq_num) <= 1) %>%
  ungroup() %>%
  select(subject_id, hadm_id, icd9_code, seq_num) %>%
  print() -> diagnose
```
Now we `inner_join` the table of admissions to ICU and the table of admissions that include MI diagnosis.
```{r}
icustays %>%
  inner_join(diagnose, by = c("subject_id", "hadm_id")) %>% 
  group_by(subject_id, hadm_id) %>%
  mutate(los = sum(los)) %>% 
  summarize_all(min, na.rm = TRUE) %>% 
  # distinct(subject_id, hadm_id, .keep_all = TRUE) %>%
  # ungroup() %>%
  print() -> icu_diagnose
```

## Transform and augment query tables

Now we create a logical variable indicating the MI is the principal diagonosis or not (according to `seq_num`).
```{r}
icu_diagnose %>%
  mutate(principal_dx = seq_num == 1) %>%
  select(-seq_num) %>%
  print() -> icu_diagnose
```

We want to add information about the severity of patients’ ailments. The `drgcodes` table contains, for `DRG` codes from the All Payers Registry (APR), severity and mortality indicators. We pull the drug severity information and right-join it to our query table.
```{r}
tbl(con, "drgcodes") %>%
  filter(str_detect(drg_type, "APR")) %>%
  select(subject_id, hadm_id, drg_severity) %>%
  right_join(icu_diagnose, by = c("subject_id", "hadm_id")) %>%
  distinct() %>% 
  mutate(drg_severity = ifelse(is.na(drg_severity), 1, drg_severity)) %>%
  print() -> icu_diagnose
```

Pull the admission time `admittime`, discharge time `dischtime`, date of birth `dob`, and date of death `dod`.
```{r}
icu_diagnose %>%
  left_join(
    select(tbl(con, "admissions"),
           subject_id, hadm_id, admittime, dischtime
    ), by = c("subject_id", "hadm_id")
  ) %>%
  left_join(
    select(tbl(con, "patients"), subject_id, dob, dod),
    by = "subject_id"
  ) %>%
  print(width = Inf) -> icu_diagnose
```

To add `age` (at admission) variable into the table. [The documentation for the patients table](https://mimic.physionet.org/mimictables/patients/) explains that patients of 90 years and older had their ages artificially inflated, so we remove these patients from the analysis.
```{r}
icu_diagnose %>%
  mutate(age = date_part("year", admittime) - date_part("year", dob)) %>%
  filter(age < 90) %>%
  mutate(age = age - ifelse(
    date_part("month", admittime) < date_part("month", dob) |
      (
        date_part("month", admittime) == date_part("month", dob) &
          date_part("day", admittime) < date_part("day", dob)
      ),
    1,
    0
  )) %>%
  select(-admittime, -dischtime, -dob, -dod) %>%
  select(subject_id, hadm_id, age, everything()) %>%
  print() -> icu_diagnose
```

Finally, let's merge some demographic information (ethnicity) into our study `icu_diagnose`.
```{r}
tbl(con, "admissions") %>%
  select(subject_id, ethnicity) %>%
  distinct() %>%
  print() -> study_subjects
```
```{r}
tbl(con, "patients") %>%
  select(subject_id, gender) %>%
  distinct() %>%
  full_join(study_subjects, by = "subject_id") %>%
  print() -> study_subjects
```
```{r}
study_subjects %>%
  semi_join(icu_diagnose, by = "subject_id") %>%
  print() -> study_subjects
```
Let's resolves ome diversity and inconsistency in the `ethnicity` field:
```{r}
unknown_ethnicity <- c(
  "OTHER",
  "UNABLE TO OBTAIN",
  "UNKNOWN/NOT SPECIFIED",
  "MULTI RACE ETHNICITY",
  "PATIENT DECLINED TO ANSWER",
  "UNKNOWN"
)

study_subjects %>%
  collect() %>%
  mutate(ethnic_group = case_when(
    str_detect(ethnicity, "^ASIAN") ~ "ASIAN",
    str_detect(ethnicity, "^BLACK") ~ "BLACK",
    str_detect(ethnicity, "^HISPANIC") ~ "HISPANIC",
    str_detect(ethnicity, "^WHITE") ~ "WHITE",
    ethnicity %in% unknown_ethnicity ~ NA_character_,
    TRUE ~ NA_character_
  )) %>%
  select(subject_id, gender, ethnic_group) %>%
  print() -> study_subjects
```

Some patients are coded as belonging to more than one ethnic group. To resolve these inconsistencies, we define a helper function to pick the modal value from a vector of values in R, which can be used by the `summarize()` function to choose one ethnic group for each patient.
```{r}
most <- function(x) {
  if (all(is.na(x))) return(NA_character_)
  y <- table(x, useNA = "no")
  if (length(which(y == max(y))) > 1) return(NA_character_)
  return(names(y)[which.max(y)])
}

study_subjects %>%
  group_by(subject_id) %>%
  summarize(ethnic_group = most(ethnic_group)) %>%
  ungroup() %>%
  mutate(ethnic_group = ifelse(is.na(ethnic_group), 
                               "UNKNOWN", ethnic_group)) %>%
  print() -> subject_ethnic_groups
```
```{r}
study_subjects %>%
  select(subject_id, gender) %>%
  left_join(subject_ethnic_groups, by = "subject_id") %>%
  print() -> study_subjects
```
Now we add the demographic information `gender` and `ethnicity` into our `icu_diagnose` table:
```{r}
icu_diagnose %>%
  left_join(study_subjects, by = "subject_id", copy = TRUE) %>%
  print() -> icu_diagnose
```

Finally, we convert the length of patient's stay in ICU into whether the patient stay more than 5 days in ICU, and get the final version of `icu_diagnose` table.

```{r}
icu_diagnose %>%
  mutate(los = (los >= 5)) %>% 
  print(width = Inf) -> icu_diagnose
icu_diagnose %>% as.tibble()
```

## CONSORT Flow Diagrams

CONSORT Flow Diagrams can be used to plot the flow of data selection of a patient cohort.   
For more details, see:
[The CONSORT Flow Diagram](http://www.consort-statement.org/consort-statement/flow-diagram). Following code shows an example. 

```{r plot}
library(shape)
library(diagram)

# set margins and multiplot
par(mfrow = c(1, 1))
par(mar = c(0, 0, 0, 0))

# initialise a plot device
openplotmat()

# position of boxes
# 1st column indicates x axis position between 0 and 1
# 2nd column indicates y axis position between 0 and 1
# automatically assigns vertical position
num_of_boxes <- 5
auto_coords = coordinates(num_of_boxes)
vert_pos = rev(auto_coords[,1])
box_pos <- matrix(nrow = num_of_boxes, ncol = 2, data = 0)
box_pos[1,] = c(0.20, vert_pos[1]) # 1st box
box_pos[2,] = c(0.70, vert_pos[2]) # 2nd box
box_pos[3,] = c(0.70, vert_pos[3]) # 3rd box
box_pos[4,] = c(0.70, vert_pos[4]) # etc...
box_pos[5,] = c(0.20, vert_pos[5])

# content of boxes
box_content <- matrix(nrow = num_of_boxes, ncol = 1, data = 0)
box_content[1] = "All admissions in MIMIC-III \n n = 58,976" # 1st box
box_content[2] = "Exclude patients with no ICU admissions \n n = 1,071" # etc...
box_content[3] = "Exclude patients with no pregancy \n n = 57260"
box_content[4] = "Exclude patients with uncentral diagnose \n n = 491"
box_content[5] = "Study cohort \n n = 154"

# adjust the size of boxes to fit content
box_x <- c(0.20, 0.25, 0.25, 0.25, 0.20)
box_y <- c(0.07, 0.07, 0.07, 0.07, 0.07)

# Draw the arrows
straightarrow(from = c(box_pos[1,1],box_pos[2,2]), to = box_pos[2,], lwd = 1)  
straightarrow(from = c(box_pos[1,1],box_pos[3,2]), to = box_pos[3,], lwd = 1)  
straightarrow(from = c(box_pos[1,1],box_pos[4,2]), to = box_pos[4,], lwd = 1)  
straightarrow(from = box_pos[1,], to = box_pos[5,], lwd = 1)

# Draw the boxes
for (i in 1:num_of_boxes) {
  textrect(mid = box_pos[i,], radx = box_x[i], rady = box_y[i], 
           lab = box_content[i], 
           shadow.col = "grey")
  }
```


# Data visualization

Now we have a study cohort consisting of pregant patient information. Some patients stay in ICU for more than 5 days and the others do not. Plot each column of the patient information with respect to whether the patient stay in ICU for more than 5 days to get the sense of inherent relationships between them.

Based on each cohort record, the age at admission is treated as a discrete variable. There are similar amounts of patients at full age that stay in ICU for more than 5 days.

```{r}
icu_diagnose %>%
  ggplot() + 
  geom_bar(mapping = aes(x = age, fill = los))
```

Based on each cohort record, the severity of patients' ailments is treated as a discrete variable. Patients with severer ailments tend to stay more than 5 days in ICU.

```{r}
icu_diagnose %>%
  ggplot() + 
  geom_bar(mapping = aes(x = drg_severity, fill = los))
```

Based on each cohort record, the ICD-9 code related to complications of pregancy, childbirth, and the puerperium is treated as a discrete variable. Patients with different ICD-9 code have significant difference in whether the length of stay in ICU is more than 5 days or not.

```{r}
icu_diagnose %>%
  ggplot() + 
  geom_bar(mapping = aes(x = icd9_code, fill = los)) + 
  theme(axis.text.x = element_text(angle = 90))
```

Based on each cohort record, whether principal diagonosis or not is treated as a discrete variable. Patients diagonosed pregancy principally tend to stay more than 5 days in ICU.

```{r}
icu_diagnose %>%
  ggplot() + 
  geom_bar(mapping = aes(x = principal_dx, fill = los))
```

Based on each cohort record, the ethnicity of patients is treated as a discrete variable. Patients with hispanic ethnicity tend to stay less than 5 days in ICU.

```{r}
icu_diagnose %>%
  ggplot() + 
  geom_bar(mapping = aes(x = ethnic_group, fill = los)) + 
  theme(axis.text.x = element_text(angle = 90))
```

# Analytics

Let's further manipulate the study cohort such that all the variables are type of numerial values between 0 and 1.

```{r}
icu_diagnose %>% 
  mutate(los = ifelse(los, 1, 0)) %>% 
  mutate(principal_dx = ifelse(principal_dx, 1, 0)) %>% 
  mutate(ethnic_group = case_when(
    str_detect(ethnic_group, "ASIAN") ~ 1, 
    str_detect(ethnic_group, "BLACK") ~ 2, 
    str_detect(ethnic_group, "HISPANIC") ~ 3, 
    str_detect(ethnic_group, "WHITE") ~ 4, 
    str_detect(ethnic_group, "UNKNOWN") ~ 5
  )) %>% 
  select(age, drg_severity, icd9_code, principal_dx, ethnic_group, los) %>% 
  mutate(age = age / 45.0) %>%
  mutate(drg_severity = drg_severity / 4) %>%
  mutate(icd9_code = as.numeric(icd9_code) / 68000) %>%
  mutate(ethnic_group = ethnic_group / 5) %>%
  print() -> raw_data0

raw_data <- as.data.frame(raw_data0) %>% 
  unname() %>% 
  # as.matrix() %>% 
  print()

x_train <- raw_data[1:130, 1:5] %>% print()
y_train <- raw_data[1:130, 6] %>% print()
dim(x_train)
dim(y_train)

x_test <- raw_data[131:154, 1:5] %>% print()
y_test <- raw_data[131:154, 6] %>% print()
dim(x_test)
dim(y_test)
# raw_data <- as.vector(raw_data) %>% print()
```

## MNIST-MLP model

Define a sequential model (a linear stack of layers) with 2 fully-connected hidden layers (50 and 25 neurons).

```{r}
library(keras)
# library(tensorflow)
```
```{r}
x_train1 <- x_train %>% as.matrix()
y_train1 <- y_train %>% as.matrix()

x_test1 <- x_test %>% as.matrix()
y_test1 <- y_test %>% as.matrix()
```

```{r}
model1 <- keras_model_sequential()
model1 %>%
  layer_dense(units = 50, activation = 'relu', input_shape = c(5)) %>%
  layer_dropout(rate= 0.4) %>%
  layer_dense(units = 25, activation = 'relu') %>%
  layer_dropout(rate = 0.3) %>%
  layer_dense(units = 5, activation = 'softmax')
summary(model1)
```

Compile the model with appropriate loss function, optimizer, and metrics.

```{r}
model1 %>% compile(
  loss = 'sparse_categorical_crossentropy',
  optimizer = optimizer_rmsprop(),
  metrics = c('accuracy')
)
```

Trainning and validation.

```{r}
system.time({
  history <- model1 %>% fit(
    x_train1, y_train1,
    epochs = 30, batch_size = 20,
    validation_split = 0.2
  )
})
```
```{r}
plot(history)
```

Evaluate model performance on the test data.

```{r}
model1 %>% evaluate(x_test1, y_test1)
```

After changing the parameter of the model for several times, loss of around 0.54 and accuracy of around 0.79 are achieved with relatively low calculation cost.

## Logistic Regression

Data preparation for logistic regression.

```{r}
library(aod)
library(ggplot2)
```
```{r}
raw_data <- as.data.frame(raw_data0) %>% 
  select(los, age, drg_severity, icd9_code, principal_dx, ethnic_group) %>% 
  print()
train <- raw_data[1:130, ] %>% as.tibble()%>% print()
test <- raw_data[131:154, ] %>% as.tibble() %>% print()
#   merge(y_train, x_train)
# names(train)[1] <- "los"
# train %>% print()
#   merge(y_test, x_test) 
# names(test)[1] <- "los"
# test %>% print()
```
```{r}
summary(train)
sapply(train, sd)
```

Define a model with logistic regression.

```{r}
# train$pricipal_dx <- factor(train$pricipal_dx)
model2 <- glm(los ~ age + drg_severity + icd9_code + principal_dx 
              + ethnic_group, data = train, family = "binomial")
summary(model2)
```

Predict the probability that the patient stays in ICU for more than 5 days with test data.

```{r}
x_test2 <- test %>% 
  select(age, drg_severity, icd9_code, principal_dx, ethnic_group)
x_test2$P <- predict(model2, newdata = x_test2, type = "response")
x_test2
```

Calculate the loss of prediction and plot.

```{r}
y_test2 <- test %>% 
  select(los)
names(y_test2)[1] <- "y"
y_predict2 <- x_test2 %>% 
  select(P)
names(y_predict2)[1] <- "y"
y <- bind_rows(y_test2, y_predict2)
type1 <- rep(c("test"), 24) %>% as.tibble()
names(type1)[1] <- "type" 
type2 <- rep(c("predict"), 24) %>% as.tibble()
names(type2)[1] <- "type" 
type <- bind_rows(type1, type2)
eva <- data.frame(rep(seq(from = 1, to = 24), 2), y, type)
names(eva)[1] <- "ind"
# names(eva)[2] <- "y"
# names(eva)[3] <- "type"
eva %>% print()
eva %>% 
  group_by(type) %>% 
  ggplot() +
  geom_line(mapping = aes(x = ind, y = y, color = type), size = 0.5) +
  scale_color_manual(values = c("blue", "black"))

# loss <- sqrt(sum((y_test2 - y_predict2)^2) / 24)
loss <- (-1) * sum(y_test2 * log(y_predict2)) / 24
loss
```

Measure how well the model fits by comparing the model with a null model.

```{r}
with(model2, null.deviance - deviance)
with(model2, df.null - df.residual)
with(model2, pchisq(null.deviance - deviance, df.null - df.residual, 
                    lower.tail = FALSE))
```

The chi-square of 7.96 with 5 degrees of freedom and an associated p-value of 0.16 tells us that the model as a whole fits better than an empty model, but it is still not accurate enough.

The loss of the logistic regression model seems to have lower loss, but since the number of samples is not abundant, and according to the plot shown above, logistic regression gives moderate predict, which actually may give subtle difference among different patients, then this low loss may not be typical.

## Close the connection to a database

Close the connection:
```{r}
dbDisconnect(con)
```

# Conclusions










