---
title: "Biostat 203B Homework 2"
author: Rui Xu
subtitle: Due Feb 7 @ 11:59PM
output:
  # ioslides_presentation: default
  html_document:
      toc: true
      toc_depth: 4
---

```{r setup, include=FALSE}
    knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

```{r}
    sessionInfo()
```

Use tidyverse (ggpot2, dplyr) to explore the [MIMIC-III](https://mimic.physionet.org) data introduced in [homework 1](https://ucla-biostat203b-2020winter.github.io/hw/hw1/hw1.html).

## Q1

Demographic information of the patients admitted into hospital is available in `ADMISSION.csv`. See <https://mimic.physionet.org/mimictables/admissions/> for details of each field in this file. Summarize following variables using appropriate graphs:   

  - admission year  
- admission month  
- admission week day  
- admission hour  
- length of hospital stay  
- admission type  
- number of admissions per patient  
- admission location  
- insurance  
- language  
- religion  
- martial status  
- ethnicity  
- death 

Note it is possible that one patient (uniquely identified by the `SUBJECT_ID`) is admitted into hospital multiple times. When summarizing some demographic information, it makes sense to summarize based on only unique patients. 

**solution:**

### Prerequisites

Load tidyverse and lubridate

```{r}
    library(tidyverse)
    library(lubridate)
```

### Import data

Let's turn flat files into tibble:

```{r}
    admission <- read_csv("/home/203bdata/mimic-iii/ADMISSIONS.csv",
                     col_types = cols(ROW_ID = col_integer(),
                                      SUBJECT_ID = col_integer(),
                                      HADM_ID = col_integer(),
                                      HOSPITAL_EXPIRE_FLAG = col_logical(),
                                      HAS_CHARTEVENTS_DATA = col_logical()))
    admission %>% print(width = Inf)
```

### Graphs

#### admission year

Based on each admission record, the admission year is treated as a continuous variable. In general, the amount of patients admitted every year does not have much difference.
 
```{r}
    admission %>%
      ggplot(aes(x = ADMITTIME)) +
      geom_freqpoly(binwidth = 31536000) + # 86400 seconds = 1 day
      labs(x = "Admission year")
```
    
#### admission month

Based on each admission record, the admission month is treated as a discrete variable. In general, the amount of patients admitted every month does not have much difference.

```{r}
    admission %>%
      mutate(adm_month = month(ADMITTIME, label = TRUE)) %>%
      ggplot(aes(x = adm_month)) +
      geom_bar() +
      labs(x = "Admission month")
```

#### admission weekday

Based on each admission record, the admission weekday is treated as a discrete variable. In general, the hospital is busier during workday than during weekend.

```{r}
    admission %>%
      mutate(adm_wday = wday(ADMITTIME, label = TRUE)) %>%
      ggplot(aes(x = adm_wday)) +
      geom_bar() +
      labs(x = "Admission weekday")
```

#### admission hour

Based on each admission record, the admission hour is treated as a discrete variable. In general, the admission hour has a peak value around 7 am and the hosiptal is busier during afternoon.

```{r}
    admission %>%
      mutate(adm_hour = hour(ADMITTIME)) %>%
      ggplot(aes(x = adm_hour)) +
      geom_bar() +
      labs(x = "Admission hour")
```
    
#### length of hospital stay

Based on each admission record, the length of hospital stay is treated as a discrete variable. In general, most of the patients stay in the hospital less than 25 days.

```{r}
    admission %>%
    mutate(los_days = as.numeric(as.duration(DISCHTIME - ADMITTIME) 
                                 / 86400)) %>%
      ggplot(aes(x = los_days)) +
      geom_histogram(bins = 100) +
      labs(x = "Length of hospital stay (days)")
```
    
#### admission type

Based on each admission record, the admission type is treated as a discrete variable. The most admission type is emergency and the least admission type is urgent.

```{r}
    adm_type <- admission %>%
      group_by(ADMISSION_TYPE) %>%
      summarise(
        count = n()
      ) %>% as.data.frame
    label_value <- paste('(', round(adm_type$count / sum(adm_type$count) * 100,
                                    1), '%)', sep = '')
    label <- paste(adm_type$ADMISSION_TYPE, label_value, sep = '')
    admission %>%
      ggplot(mapping = aes(x = 'portion', fill = ADMISSION_TYPE)) +
      geom_bar(width = 1) +
      coord_polar("y") +
      labs(x = '', y = '', title = '') +
      theme(axis.text = element_blank()) +
      theme(axis.ticks = element_blank()) +
      scale_fill_discrete(labels = label)
```

#### number of admissions per patient

Based on each unique patient, the number of admissions per patient is treated as a discrete variable. Most of the patients are only admitted to the hospital once.

```{r}
    adm_num <- admission %>%
       group_by(SUBJECT_ID) %>%
       summarise(
         count = n()
       )
    adm_num %>%
      ggplot(aes(x = count)) +
      geom_bar() +
      labs(x = "number of admissions per patient")
```

#### admission location

Based on each admission record, the admission location is treated as a discrete variable. The most admission location is emergency room admit, which corresponds to the most admission type.

```{r}
    adm_loc1 <- admission %>%
      group_by(ADMISSION_LOCATION) %>%
      summarise(
        count = n()
      ) %>%
      filter(count > 600)
    adm_loc2 <- admission %>%
      group_by(ADMISSION_LOCATION) %>%
      summarise(count = n()) %>%
      filter(count <= 600) %>%
      mutate(tmp = sum(count))
    adm_loc3 <- tribble(~ADMISSION_LOCATION, ~count,
                        "OTHER", adm_loc2$tmp[1])
    adm_loc <- bind_rows(adm_loc1, adm_loc3) %>% as.data.frame
    label_value <- paste('(', round(adm_loc$count / sum(adm_loc$count) * 100,
                                    1), '%)', sep = '')
    label <- paste(adm_loc$ADMISSION_LOCATION, label_value, sep = '')
    adm_loc %>%
      ggplot(mapping = aes(x = 'portion', fill = ADMISSION_LOCATION)) +
      geom_bar(width = 1) +
      coord_polar("y") +
      labs(x = '', y = '', title = '') +
      theme(axis.text = element_blank()) +
      theme(axis.ticks = element_blank()) +
      scale_fill_discrete(labels = label)
```

#### insurance

Based on each unique patient, we can get a sense of the insurance distribution. The insurance is treated as a discrete variable. Most of the patients use medicare and least of the patients use pay by themselves.

```{r}
    people <- distinct(admission, SUBJECT_ID, .keep_all = TRUE)
    insurance <- people %>%
      group_by(INSURANCE) %>%
      summarise(
        count = n()
      ) %>% as.data.frame
    label_value <- paste('(', round(insurance$count / sum(insurance$count)
                                    * 100, 1), '%)', sep = '')
    label <- paste(insurance$INSURANCE, label_value, sep = '')
    people %>%
      ggplot(mapping = aes(x = 'portion', fill = INSURANCE)) +
      geom_bar(width = 1) +
      coord_polar("y") +
      labs(x = '', y = '', title = '') +
      theme(axis.text = element_blank()) +
      theme(axis.ticks = element_blank()) +
      scale_fill_discrete(labels = label)
```

#### language

Based on each unique patient, we can get a sense of the language distribution. The language is treated as a discrete variable. Most of the patients speak English.

```{r}
    people <- distinct(admission, SUBJECT_ID, .keep_all = TRUE)
    language1 <- people %>%
      group_by(LANGUAGE) %>%
      summarise(
        count = n()
      ) %>%
      filter(count > 460)
    language2 <- people %>%
      group_by(LANGUAGE) %>%
      summarise(count = n()) %>%
      filter(count <= 460) %>%
      mutate(tmp = sum(count))
    language3 <- tribble(~LANGUAGE, ~count,
                        "OTHER", language2$tmp[1])
    language <- bind_rows(language1, language3) %>% as.data.frame
    label_value <- paste('(', round(language$count / sum(language$count)
                                    * 100, 1), '%)', sep = '')
    label <- paste(language$LANGUAGE, label_value, sep = '')
    language %>%
      ggplot(mapping = aes(x = 'portion', fill = LANGUAGE)) +
      geom_bar(width = 1) +
      coord_polar("y") +
      labs(x = '', y = '', title = '') +
      theme(axis.text = element_blank()) +
      theme(axis.ticks = element_blank()) +
      scale_fill_discrete(labels = label)
```

#### religion

Based on each unique patient, we can get a sense of the religion distribution. The religion is treated as a discrete variable. Most of the patients are catholic.

```{r}
    people <- distinct(admission, SUBJECT_ID, .keep_all = TRUE)
    religion1 <- people %>%
      group_by(RELIGION) %>%
      summarise(
        count = n()
      ) %>%
      filter(count > 460, RELIGION != "OTHER")
    religion2 <- people %>%
      group_by(RELIGION) %>%
      summarise(count = n()) %>%
      filter(count <= 460 | RELIGION == "OTHER") %>%
      mutate(tmp = sum(count))
    religion3 <- tribble(~RELIGION, ~count,
                        "OTHER", religion2$tmp[1])
    religion <- bind_rows(religion1, religion3) %>% as.data.frame
    label_value <- paste('(', round(religion$count / sum(religion$count)
                                    * 100, 1), '%)', sep = '')
    label <- paste(religion$RELIGION, label_value, sep = '')
    religion %>%
      ggplot(mapping = aes(x = 'portion', fill = RELIGION)) +
      geom_bar(width = 1) +
      coord_polar("y") +
      labs(x = '', y = '', title = '') +
      theme(axis.text = element_blank()) +
      theme(axis.ticks = element_blank()) +
      scale_fill_discrete(labels = label)
```

#### martial status

Based on each unique patient, we can get a sense of the martial status distribution. The martial status is treated as a discrete variable. Most of the patients are married.

```{r}
    people <- distinct(admission, SUBJECT_ID, .keep_all = TRUE)
    martst <- people %>%
      group_by(MARITAL_STATUS) %>%
      summarise(
        count = n()
      ) %>% as.data.frame
    label_value <- paste('(', round(martst$count / sum(martst$count)
                                    * 100, 1), '%)', sep = '')
    label <- paste(martst$MARITAL_STATUS, label_value, sep = '')
    people %>%
      ggplot(mapping = aes(x = 'portion', fill = MARITAL_STATUS)) +
      geom_bar(width = 1) +
      coord_polar("y") +
      labs(x = '', y = '', title = '') +
      theme(axis.text = element_blank()) +
      theme(axis.ticks = element_blank()) +
      scale_fill_discrete(labels = label)
```

#### ethnicity

Based on each unique patient, we can get a sense of the ethnicity distribution. The ethnicity is treated as a discrete variable. Most of the patients are white.

```{r}
    people <- distinct(admission, SUBJECT_ID, .keep_all = TRUE)
    enth1 <- people %>%
      group_by(ETHNICITY) %>%
      summarise(
        count = n()
      ) %>%
      filter(count > 460)
    enth2 <- people %>%
      group_by(ETHNICITY) %>%
      summarise(count = n()) %>%
      filter(count <= 460) %>%
      mutate(tmp = sum(count))
    enth3 <- tribble(~ETHNICITY, ~count,
                        "OTHER", enth2$tmp[1])
    enth <- bind_rows(enth1, enth3) %>% as.data.frame
    label_value <- paste('(', round(enth$count / sum(enth$count)
                                    * 100, 1), '%)', sep = '')
    label <- paste(enth$ETHNICITY, label_value, sep = '')
    enth %>%
      ggplot(mapping = aes(x = 'portion', fill = ETHNICITY)) +
      geom_bar(width = 1) +
      coord_polar("y") +
      labs(x = '', y = '', title = '') +
      theme(axis.text = element_blank()) +
      theme(axis.ticks = element_blank()) +
      scale_fill_discrete(labels = label)
```

#### death

Since the death analysis for each admission makes more sense than the death analysis for each person, the death analysis is based on each admission record. After plotting the death time with respect to the year, month, weekday and hour, we can conclude that the death happens randomly everyday but most of the death happens in the noon.

```{r}
    admission %>%
      filter(!is.na(DEATHTIME)) %>%
      ggplot(aes(x = DEATHTIME)) +
      geom_freqpoly(binwidth = 31536000) +
      labs(x = "Death year")
```

```{r}
    admission %>%
      filter(!is.na(DEATHTIME)) %>%
      mutate(death_month = month(DEATHTIME, label = TRUE)) %>%
      ggplot(aes(x = death_month)) +
      geom_bar() +
      labs(x = "Death month")
```

```{r}
    admission %>%
      filter(!is.na(DEATHTIME)) %>%
      mutate(death_wday = wday(DEATHTIME, label = TRUE)) %>%
      ggplot(aes(x = death_wday)) +
      geom_bar() +
      labs(x = "Death weekday")
```

```{r}
    admission %>%
      filter(!is.na(DEATHTIME)) %>%
      mutate(death_hour = hour(DEATHTIME)) %>%
      ggplot(aes(x = death_hour)) +
      geom_bar() +
      labs(x = "Death hour")
```

## Q2

Link the data in `ADMISSION.csv` and `PATIENTS.csv` (<https://mimic.physionet.org/mimictables/patients/>) and summarize following variables using appropriate graphs:  

- gender  
- age at admission 

**solution:**

### Import data and link the data

Let's turn flat files into tibbles and link the tibbles. `PATIENTS.csv` contains unique patient information. Then through left joint to admission information, every admission record is linked to one patient while the patient information shows up multiple times.

```{r}
    admission <- read_csv("/home/203bdata/mimic-iii/ADMISSIONS.csv",
                         col_types = cols(ROW_ID = col_integer(),
                                          SUBJECT_ID = col_integer(),
                                          HADM_ID = col_integer(),
                                          HOSPITAL_EXPIRE_FLAG = col_logical(),
                                          HAS_CHARTEVENTS_DATA = col_logical()))
    patient <- read_csv("/home/203bdata/mimic-iii/PATIENTS.csv")
    adm_pinfo <- left_join(admission, patient, by = "SUBJECT_ID")
    adm_pinfo %>% print(width = Inf)
```

### Graphs

#### gender

Since the gender analysis for each admission makes more sense than the gender analysis for each person, the gender analysis is based on each admission record. The ratio of male patients to female patients is around 1 and the amount of male patients is larger than the amount of female patients.

```{r}
    adm_gender <- adm_pinfo %>%
      group_by(GENDER) %>%
      summarise(
        count = n()
      ) %>% as.data.frame
    label_value <- paste('(', round(adm_gender$count / sum(adm_gender$count) * 100,
                                    1), '%)', sep = '')
    label <- paste(adm_gender$GENDER, label_value, sep = '')
    adm_pinfo %>%
      ggplot(mapping = aes(x = 'portion', fill = GENDER)) +
      geom_bar(width = 1) +
      coord_polar("y") +
      labs(x = '', y = '', title = '') +
      theme(axis.text = element_blank()) +
      theme(axis.ticks = element_blank()) +
      scale_fill_discrete(labels = label)
```

#### age at admission

Based on each admission record, the age at admission is treated as a continuous variable. The most age at admission is around 75 years old and the newborns also take a big part.

```{r}
    adm_pinfo %>%
      mutate(AGE = as.numeric(as.duration(ADMITTIME - DOB) / 31536000)) %>%
      ggplot(aes(x = AGE)) +
      geom_freqpoly(bins = 100) +
      labs(x = "Age at admission")
```

## Q3

`ICUSTAYS.csv` (<https://mimic.physionet.org/mimictables/icustays/>) contains data about Intensive Care Units (ICU) stays. Summarize following variables using appropriate graphs:  

- length of ICU stay  
- first ICU unit  
- gender  
- age

**solution:**

### Import data and link the data

Let's turn flat files into tibbles and link the tibbles. `PATIENTS.csv` contains unique patient information. Then through left joint to ICU stay information, every ICU admission record is linked to one patient while the patient information shows up multiple times.

```{r}
    patient <- read_csv("/home/203bdata/mimic-iii/PATIENTS.csv")
    icustays <- read_csv("/home/203bdata/mimic-iii/ICUSTAYS.csv")
    icustay <- left_join(icustays, patient, by = "SUBJECT_ID")
    icustay %>% print(width = Inf)
```

### Graphs

#### length of ICU stay

Based on each ICU admission record, the length of ICU stay is treated as a discrete variable. Most of the patients stay in the ICU for less than 12 days.

```{r}
    icustay %>%
      mutate(los_days = as.numeric(as.duration(OUTTIME - INTIME) / 86400)) %>%
      ggplot(aes(x = los_days)) +
      geom_histogram(bins = 100) +
      labs(x = "Length of ICU stay")
```

#### first ICU unit

Based on each ICU admission record, the first ICU unit is treated as a discrete variable. Most of the patients are admitted to MICU at the beginning.

```{r}
    ICUunit <- icustay %>%
      group_by(FIRST_CAREUNIT) %>%
      summarise(
        count = n()
      ) %>% as.data.frame
    label_value <- paste('(', round(ICUunit$count / sum(ICUunit$count) * 100,
                                    1), '%)', sep = '')
    label <- paste(ICUunit$FIRST_CAREUNIT, label_value, sep = '')
    icustay %>%
      ggplot(mapping = aes(x = 'portion', fill = FIRST_CAREUNIT)) +
      geom_bar(width = 1) +
      coord_polar("y") +
      labs(x = '', y = '', title = '') +
      theme(axis.text = element_blank()) +
      theme(axis.ticks = element_blank()) +
      scale_fill_discrete(labels = label)
```

#### gender

Based on each ICU admission record, the gender is treated as a discrete variable. The ratio of male patients to female patients is around 1 and the amount of male patients is larger than the amount of female patients.

```{r}
    gender <- icustay %>%
      group_by(GENDER) %>%
      summarise(
        count = n()
      ) %>% as.data.frame
    label_value <- paste('(', round(gender$count / sum(gender$count) * 100,
                                    1), '%)', sep = '')
    label <- paste(gender$GENDER, label_value, sep = '')
    icustay %>%
      ggplot(mapping = aes(x = 'portion', fill = GENDER)) +
      geom_bar(width = 1) +
      coord_polar("y") +
      labs(x = '', y = '', title = '') +
      theme(axis.text = element_blank()) +
      theme(axis.ticks = element_blank()) +
      scale_fill_discrete(labels = label)
```

#### age

Based on each ICU admission record, the age at admission is treated as a continuous variable. The most age at admission is around 75 years old and the newborns also take a big part.

```{r}
    icustay %>%
      mutate(age = as.numeric(as.duration(INTIME - DOB) / 31536000)) %>%
      ggplot(aes(x = age)) +
      geom_freqpoly(bins = 100) +
      labs(x = "Age at admission")
```
    
## Q4 

`CHARTEVENTS.csv` (<https://mimic.physionet.org/mimictables/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `ITEMID` variable indicates a single measurement type in the database. The `VALUE` variable is the value measured for `ITEMID`. 

`D_ITEMS.csv` (<https://mimic.physionet.org/mimictables/d_items/>) is the dictionary for the `ITEMID` in `CHARTEVENTS.csv`. Find potential values of `ITEMID` that correspond to systolic blood pressure, i.e., `LABEL` contains the string `systolic`. 

Compile a tibble that contains the first ICU stay of unique patients, with the patient's demographic information, the first systolic blood pressure measurement during ICU stay, and whether the patient died within 30 days of hospitcal admission.


**solution:**

### Prerequisites

Load htmlwidgets

```{r}
    library(htmlwidgets)
```

### Import data

```{r}
    admission <- read_csv("/home/203bdata/mimic-iii/ADMISSIONS.csv",
                         col_types = cols(ROW_ID = col_integer(),
                                          SUBJECT_ID = col_integer(),
                                          HADM_ID = col_integer(),
                                          HOSPITAL_EXPIRE_FLAG = col_logical(),
                                          HAS_CHARTEVENTS_DATA = col_logical()))
    patient <- read_csv("/home/203bdata/mimic-iii/PATIENTS.csv")
    icustay <- read_csv("/home/203bdata/mimic-iii/ICUSTAYS.csv")
    chart <- read_csv("/home/203bdata/mimic-iii/CHARTEVENTS.csv")
    item <- read_csv("/home/203bdata/mimic-iii/D_ITEMS.csv")
    admission %>% print(width = Inf)
    patient %>% print(width = Inf)
    icustay %>% print(width = Inf)
    chart %>% print(width = Inf)
    item %>% print(width = Inf)
```

### Compile the tibble

- Select the systolic blood pressure measurement information:

```{r}
    bloodp <- item %>%
      filter(str_detect(LABEL, "systolic")) %>%
      select(ITEMID, LABEL)
    bloodp2 <- inner_join(chart, bloodp, by = "ITEMID") %>%
      group_by(SUBJECT_ID) %>%
      filter(rank(STORETIME) == 1) %>%
      select(SUBJECT_ID, HADM_ID, LABEL)
    bloodp2 %>% print(width = Inf)
```

- select the ICU stay information:

```{r}
    first_icu <- icustay %>%
      group_by(SUBJECT_ID) %>%
      filter(rank(INTIME) == 1) %>%
      select(SUBJECT_ID, HADM_ID, FIRST_CAREUNIT)
    first_icu %>% print(width = Inf)
```

- Select the patient's demographic and death information:

```{r}
    pinfo <- admission %>%
      mutate(dead30 = ((as.numeric(as.duration(DEATHTIME - ADMITTIME) / 86400)
                                  ) <= 30)) %>%
      group_by(SUBJECT_ID) %>%
      filter(rank(desc(dead30)) == 1) %>%
      select(SUBJECT_ID, HADM_ID, LANGUAGE: ETHNICITY, dead30)
    pinfo %>% print(width = Inf)
```

- compile the tibble

```{r}
    info <- pinfo %>%
      full_join(bloodp2, by = "HADM_ID") %>%
      full_join(first_icu, by = "HADM_ID")
    info %>% print(width = Inf)
```


